Tail associations in ecological variables and their impact on extinction risk

### 2020-01-21 Response.txt

•       In the second paragraph in Discussion, X’s are used to refer to generic random variables, but it appeared rather abruptly as the letter was not used to denote variables in the paper up to that point in the main text. It would be helpful to give the definition parenthetically in the text. Do i and j refer to different weather variables, to different ecological variable, or also to different patches?

>>>We apologize for this omission! We have now added the definition. We have also clarified the rest of this paragraph.

•       The paragraph starting from Line 372 is very long and winding. This needs to be tightened substantially (to 2/3 to ½ of the current length, which is 1.5 pages).

>>>We split the paragraph into two and shortened, as the referee requests.

@@

•       Line 419-421. This sentence is very confusing. Using “/” and “(respectively)” makes it almost intractable.

>>>We revised, as requested.

•       Line 424-450. I suggest that the sentence starting with “Although” should come after “…with asymmetric tail associations” in Line 445 for a better flow. As a connection, add a sentence like “This result is partly consistent with the prediction from Tuljapurkar’s approximation” before “The long-term stochastic growth rate… (Line 431)”.

>>>In the course of performing other edits, we revised this paragraph so the comment no longer applies.

### 2020-01-21 MainText.Rmd
even closer, just one thing to add plus the new figure and related text

<!--future ideas start here-->
Our initial exploration for stage-structured models suggested limited influence of 
tail associations between vital rates on extinction risk, but some
additional lines of inquiry might be worth exploring as part of an effort to
understand how general our results are.
First, we considered a model with positive association between 
egg and adult vital rates,
corresponding to similar impacts of an environmental variable on 
eggs and adults. Negative associations between vital rates
can occur as well, and the impacts of tail associations should be considered
in that context as well. Negative associations relate to
life history trade-offs, e.g. between survival and reproduction 
(e.g, @Lande2003, p. 59). Asymmetric tail associations may occur
in such a context if, for instance, there is a three way trade-off
among vital rates $r_1$, $r_2$, and $r_3$. Values of $r_1$ near
an upper limit may be strongly associated with,
for instance, $r_2$, whereas lower values of $r_1$ may not be strongly
associated with $r_2$ because reductions/increases in $r_1$ can
then be traded off against increases/reductions in either 
$r_2$ or $r_3$. It is possible that tail associations in negatively 
related vital rates produce different outcomes for extinction risk
from what we observed by studying positively associated vital rates.

<!--Tuljapurkar approximation-->
A second topic of research that our work suggests has to do with Tuljapurkar's
well-known approximation of the long-term stochastic growth rate [@tulja1982; @tulja1990].
It may be possible to obtain a better understanding, using Tuljapurkar's
approximation, of our result that asymmetries of tail association
may have limited influence on the extinction risk of 
stage-structured, non-spatial models.
The long-term stochastic growth rate is an important quantity in demography
[@Caswell2000] and population viability analysis for conservation 
[@MorrisDoak2002], and it factors prominently into
extinction risk estimates for density-independent age- or 
stage-structured population models 
[@Lande1988], $\vec{P}(t+1)=M(t)\vec{P}(t)$. Here $\vec{P}(t)$ is the 
age- or stage-structured population and $M$ is a matrix-valued stationary stochastic 
process. The matrix entries of $M$ can correspond, for instance, to the fluctuating
fecundity and survival rates of the life stages of the population. 
For such a model, Tuljapurkar's approximation expresses the long-term stochastic 
growth rate as a function of the variances and covariances of the entries
of $M$ and of the eigenvalues and eigenvectors of the expected value of 
$M$. None of these quantities depends on tail associations between matrix entries.
Thus, whenever Tuljapurkar's approximation is a good
approximation, the long-term stochastic growth rate should not depend sensitively
on tail association. @Lande2003 provides reasoning supporting a claim that 
Tuljapurkar's approximation
is "likely to be accurate in many cases," (p. 60) though we are unaware of the extent
to which accuracy of the approximation has been tested for scenarios with
asymmetric tail associations. It seems potentially worth exploring the accuracy of 
Tuljapurkar's approximation under scenarios of tail association in the vital rates
and relating results to the effects of tail associations on the long-term
stochastic growth rate and extinction risks.
We note that the above reasoning pertains only to density-independent models,
and says nothing about our results for the density-dependent version of our
stage-structured model.

>>>>

<!--future ideas start here-->
Our initial exploration for stage-structured models suggested limited influence of 
tail associations between vital rates on extinction risk, but some
additional lines of inquiry might be worth exploring as part of an effort to
understand how general our results are.
First, we considered a model with positive association between 
egg and adult vital rates,
corresponding to similar impacts of an environmental variable on 
eggs and adults. Negative associations between vital rates
can occur as well, and the impacts of tail associations should be considered
in that context as well. Negative associations relate to
life history trade-offs, e.g. between survival and reproduction 
(e.g, @Lande2003, p. 59). Asymmetric tail associations may occur
in such a context if, for instance, there is a three way trade-off
among vital rates $r_1$, $r_2$, and $r_3$. Values of $r_1$ near
an upper limit may be strongly associated with $r_2$, whereas lower 
values of $r_1$ may not be strongly
associated with $r_2$ because $r_1$ trades off against both 
$r_2$ and $r_3$. It is possible that tail associations in negatively 
related vital rates produce different outcomes for extinction risk
from what we observed by studying positively associated vital rates.

<!--Tuljapurkar approximation-->
A second topic of research that our work suggests has to do with Tuljapurkar's
well-known approximation of the long-term stochastic growth rate [@tulja1982; @tulja1990].
It may be possible to obtain a better understanding, using Tuljapurkar's
approximation, of our result that asymmetries of tail association
may have limited influence on the extinction risk of 
stage-structured, non-spatial models.
The long-term stochastic growth rate is an important quantity in demography
[@Caswell2000] and population viability analysis for conservation 
[@MorrisDoak2002], and it factors prominently into
extinction risk estimates for density-independent age- or 
stage-structured population models 
[@Lande1988], $\vec{P}(t+1)=M(t)\vec{P}(t)$. Here $\vec{P}(t)$ is the 
age- or stage-structured population and $M$ is a matrix-valued stationary stochastic 
process. The matrix entries of $M$ can correspond, for instance, to the fluctuating
fecundity and survival rates of the life stages of the population. 
For such a model, Tuljapurkar's approximation expresses the long-term stochastic 
growth rate as a function of the variances and covariances of the entries
of $M$ and of the eigenvalues and eigenvectors of the expected value of 
$M$. None of these quantities depends on tail associations between matrix entries.
Thus, whenever Tuljapurkar's approximation is a good
approximation, the long-term stochastic growth rate should not depend sensitively
on tail association. @Lande2003 provides reasoning supporting a claim that 
Tuljapurkar's approximation
is "likely to be accurate in many cases," (p. 60) though we are unaware of the extent
to which accuracy of the approximation has been tested for scenarios with
asymmetric tail associations. It seems potentially worth exploring the accuracy of 
Tuljapurkar's approximation under scenarios of tail association in the vital rates;
and then relating results to the effects of tail associations on the long-term
stochastic growth rate and extinction risks.
We note that the above reasoning pertains only to density-independent models,
and says nothing about our results for the density-dependent version of our
stage-structured model.
